Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-06-09 pkg/paper/Paper/draft.Rnw
draft changed
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@115 b679b2f7-6539-45d0-a605-5ddcb976cde9

\Abstract{
  In this paper, we test for structural breaks in the inflation dynamics of 21 European countries. To capture
  changes in the first three moments of the inflation rates, we assumed them to be distributed according to 
  a generalized logistic distribution. We find breaks in all but 4 countries.
  Overall, there is a convergence towards a lower inflation rate, but with an increase in variance and a change
  towards negative skewness.
  %% avoid "we"-constructs
  %% "adopt" better than "assume"
}

@@

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  A-6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
}

@@

\readme{perhaps skip this paragraph?} The way towards the EMU basically consisted of three stages: stage~I
(1990--1994) as a phase of liberalization, stage~II (1995--1998) a phase of
convergence, and stage~III (1999--2001) the transition period, which ended with
the introduction of the euro as legal tender. A possible fourth stage is the
continuing integration of new member states, mostly former communist countries.

Since the beginning of the European Union, the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area \citep[as developed in:][]{Mundell:1961}, 
all the way to the very survival of the euro in light of the budgetary problems of some of its member
states. The effects of monetary unions on a number of macroeconomic indicators,
with inflation being the most important one in this context, is in the center of
an ongoing debate. This concerns the issue of short--run and steady--state
inflation uncertainty -- in context to inflation expectations -- dealt with in
\citet{cap}, or the degree of similarity of short--run dynamic properties of the inflation rates in EU
countries, which is the topic of an investigation by \citet{palomba}.

@@

@@

\fixme{comment on results here} Overall, we find four periods during which structural breaks
happened. Two of them in the 1990s (early and late), another one in the early 2000s, and the last period covers the financial crisis. There is also a 
tendency towards a lower mean inflation rate, combined with a decreasing difference between
individual inflation rates.

@@

\begin{figure}[t!]
<<slovenia-plots1, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
plot(x$data, main = "", ylab = "Monthly inflation rate (in percent)", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
if(x$nbreaks > 0) lines(confint(x$breakpoints, breaks = x$nbreaks))
plot(x$efp, functional = supLM(0.1), main = "", ylab = "LM statistics")
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{\label{fig:slovenia1} Inflation rate series for Slovenia
with breakpoint estimate (\Sexpr{breakdates(x$breakpoints)}), associated confidence interval, and fitted mean
from generalized logistic distribution (left). Corresponding sup$\mathit{LM}$ test (right)
with sequence of $\mathit{LM}$~statistics,
critical value at 5\% level (horizontal red line line), and estimated breakpoint (vertical dashed line).}

<<slovenia-plots2, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
for(i in 1:2) plot(x$model[[i]], moments = TRUE,
  xlim = c(-1.5, 2.2), ylim = c(0, 1.3),
  breaks = seq(-1.5, 2, by = 0.25),
  xlab = paste(names(x$model)[i], "\n", "Goodness-of-fit p-value: ",
    format.pval(summary(x$model[[i]])$chisq.test$p.value, digits = 4),
    sep = ""))
@
\caption{\label{fig:slovenia2} Histogramm of observed monthly inflation rates in
Slovenia for \Sexpr{names(x$model)[1]} (left) and \Sexpr{names(x$model)[1]} (right),
along with fitted generalized logistic probability density function, associated
moments, and $\chi^2$ goodness-of-fit test.}
\end{figure}

@@

To explore the question of convergence, we proceed in two steps:
First, in Section~\ref{sec:bpresults}, the estimation technique proposed
in Section~\ref{sec:model} is applied to each of the countries included in
the data set (see Section~\ref{sec:data}). Second, Section~\ref{sec:aggregate}
aggregates the results from each individual country to reveal different
potential convergence patterns over time within three groups of countries:
core-euro (\Sexpr{paste(colnames(hicps)[euro], collapse = ", ")}),
PIIGS  (\Sexpr{paste(colnames(hicps)[piigs], collapse = ", ")}), and the
rest.

@@

\begin{table}[p!]
\begin{center}
\footnotesize
\begin{tabular}{llrrrlll}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline

@@

Finally, the inflation rates for the whole euro area \fixme{Is EU euro area or European union?},
as provided by the OECD, are considered and three different regimes are found: (1)~until 
1999 with both a low mean and variance, (2)~the third and final stage of monetary integration,
accompanied with a significant increase in the mean, (3)~a period encompassing
the overheating of the economy that culminated in the financial crisis, showing a big increase
in variance and a lower mean inflation.

\fixme{some concluding remarks about the EU level bridging the current gap to the next subsection}

@@

\subsection{Convergence on an aggregate level}\label{sec:aggregate}

\begin{table}[t!]
\begin{center}
\begin{tabular}{llrrrr}
\hline
Moment & Quartile & \Sexpr{yrs[1]} & \Sexpr{yrs[2]} & \Sexpr{yrs[3]} & \Sexpr{yrs[4]}\\
\hline

@@

\fixme{also ``zoom out'' more, a big picture needs to be established}

Table~\ref{tab:results2} presents lower quartile,
median, and upper quartile (25\%, 50\%, 75\% percentiles) of the first three moments of the
inflation series in January \Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]} and \Sexpr{yrs[4]}. The countries are grouped
into the ``core'' of the eurozone, the PIIGS countries  and all the others. For this, we took account of possible
breaks in the moments.

The time periods are chosen according to four fundamental change periods. Namely, the early 1990s, with breaks in
Austria, Finland, France, Sweden, and UK. The late 1990s, with breaks in Greece, Italy, Luxembourg, Portugal, Czech Republic,
Estonia, and Hungary. And also the period after the second phase of convergence, with breaks in the early 2000s (Poland, Slovakia, and Slovenia)
and the two countries with the most visible effect of the financial crisis (Ireland and Spain).

In the core-zone, there is a decrease in the lower quartile from \Sexpr{yrs[1]} to
\Sexpr{yrs[2]}. However, the interquartile range decreased by almost 20\%. The variance increased
slightly, but also with a 15\% decrease in interquartile range. The skewness
stayed symmetric in the median, with almost an 8\% increase in the interquartile range.

The PIIGS countries significantly decreased the median inflation rates from \Sexpr{yrs[1]} to \Sexpr{yrs[4]}. 
Whereas the interquartile range was very tight in \Sexpr{yrs[1]}, it increased in \Sexpr{yrs[2]} and again in \Sexpr{yrs[4]}.
However, the median values are much smaller. The variance increased from \Sexpr{yrs[1]} to \Sexpr{yrs[2]} but did not change much 
in \Sexpr{yrs[4]}. The interquartile range did not change much too. Median skewness changed considerable 
from highly positive to negative values. This is mostly due to the changes associated with the financial crisis, that was
responsible for deflationary shocks. The interquartile range of skewness decreased considerably.

The other countries were successful in decreasing all three moments from \Sexpr{yrs[1]} to \Sexpr{yrs[3]}. Still, their median
inflation rate is higher than those of the core countries. Also, the interquartile range of the mean decreased considerably. 
Contrary to the euro--zone and the PIIGS countries, variance decreased also, both in terms of the median value and the
interquartile range. The skewness decreased as well, as did the interquartile range of the skewness parameter.

In context of the convergence issue, we find that, with the interquartile range of the mean of all the aggregated zones declining and getting tighter, there is  
evidence of a convergence towards some common mean inflation rate in the countries of our sample, both within the subgroups and the three groups put together.
This is also visible in the median values, that  diverge much less than at the beginning in \Sexpr{yrs[1]}.

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\begin{center}
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  gray_red1 <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0))
  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
    xlab = "Time", ylab = "Monthly inflation rates")
  for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (gray) and estimated means from
12-month rolling GL fits (red) over time for all countries.}
\end{center}
\end{figure}
\setkeys{Gin}{width=\textwidth}

@@

More specifically, the mean inflation of the euro area \fixme{see above} from a GL fit in the 
year before the crisis (\Sexpr{yearmon(2007 + 7/12)}--\Sexpr{yearmon(2008 + 6/12)}) was
$\Sexpr{round3(glmean(window(eu, start = yearmon(2007 + 7/12), end = yearmon(2008 + 6/12))))}$
and after the crisis (\Sexpr{yearmon(2008 + 7/12)}--\Sexpr{yearmon(2009 + 6/12)}) it dropped to
$\Sexpr{round3(glmean(window(eu, start = yearmon(2008 + 7/12), end = yearmon(2009 + 6/12))))}$
before rising again in the following year  (\Sexpr{yearmon(2009 + 7/12)}--\Sexpr{yearmon(2010 + 6/12)}):
$\Sexpr{round3(glmean(window(eu, start = yearmon(2009 + 7/12), end = yearmon(2010 + 6/12))))}$.
Another way to bring out this pattern in a graphical way is
Figure~\ref{fig:glmean}. It depicts the inflation rates (in gray) and a 12-month rolling mean (in red),
calculated again as the first moment of a fitted GL distribution.
The strong increase in mean inflation during the boom years of 2007/2008
accompanied by an equally strong deflation following the burst of the bubble
is clearly visible for most countries. However, the level of mean inflation
did not change persistently, as the inflation rates seem to have returned to
their prior path, conveying the same message as the inflation series of the
euro area.

@@

\readme{start with 1) short recap of what we did, 2) conclusions}

To assess changes in the dynamics of the inflation rates of a number of European countries both within
and without the EMU, we developed a new method for estimating structural breaks resting on the 
assumption that the inflation rates are GL distributed.

The results suggest that a GL distribution is appropriate for the modeling of the seasonally 
adjusted inflation rate series and that failing to account for a change in the skewness of the 
distribution might lead to wrong conclusions concerning the dynamics.

The results of \citet{holte}, whose simulations indicated that the standard deviation of the home 
CPI inflation rate can be substantially reduced by joining a monetary union, can not be empirically validated.

For the countries in the core of the eurozone, one clearly visible trend is the rise in 
inflation volatility in almost half of the euro countries. \citet{emerson} emphasize  the fact that a high inflation rate is also more
variable and uncertain and thus causes more relative price variability, leading
to a less efficient price mechanism. From an economic point of view, a very volatile inflation rate will 
likely  contribute to greater macroeconomic instability. In the literature, there is no consensus about the effects of higher 
inflation volatility on the inflation rate. \citet{grier} find no evidence 
that higher inflation uncertainty raises the average inflation rate -- at least in the case of the USA. However,
\cite{jaro} finds a positive correlation between the level and the standard deviation of inflation. Naturally, an
efficient price mechanism will benefit from inflation rates that do not diverge too much from the expectations.

The good news is that mean inflation declined considerably, most notably in the Eastern European countries that clearly 
benefited from the monetary discipline that is demanded for those countries who want to join the EMU.

@@

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p!]
\begin{center}
<<graphs-nochange, fig=TRUE, height=6.3, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(2, 4, 8, 14)))
@
\caption{\label{fig:nochange} No change.}

<<graphs-phase1, fig=TRUE, height=4.9, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(1, 6, 7)))
@
\caption{\label{fig:phase1} Phase~1 convergence.}
\end{center}
\end{figure}

\begin{figure}[p!]
\begin{center}
<<graphs-phase2, fig=TRUE, height=6.3, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(9, 12, 13, 16)))
@
\caption{\label{fig:phase2} Phase~2 convergence.}

<<graphs-crisis, fig=TRUE, height=3.5, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(11, 19)))
@
\caption{\label{fig:crisis} Financial crisis.}
\end{center}
\end{figure}

\begin{figure}[p!]
\begin{center}
<<graphs-noneuro, fig=TRUE, height=3.5, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(20, 21)))
@
\caption{\label{fig:noneuro} Non-euro.}

<<graphs-eastern, fig=TRUE, height=9.1, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(3, 5, 10, 15, 17, 18)))
@
\caption{\label{fig:eastern} Eastern countries.}
\end{center}
\end{figure}

>>>>

\renewcommand{\abstractname}{Summary}
\Abstract{
	In an attempt to assess the effects of the EMU on inflation rate dynamics of its member states, this paper tests for structural breaks in the inflation dynamics of 21 European   countries. To capture   changes in the first three moments of the inflation rates, a model is adopted that makes use of the generalized logistic distribution. 
	There are   breaks in all but 4 countries.  Also, the differences between the inflation dynamics and its changes  are considerable. 
  Overall, there is a convergence towards a lower inflation rate, but with an increase in variance and a decrease in skewness.
  %% avoid "we"-constructs
  %% "adopt" better than "assume"
}

@@

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  A-6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
  Telephone: +43/ 512/ 507 - 7103 \\
  Fax: 43/ 512/ 507 - 2851 \\
}

@@

\fixme{affiliation should appear on title page}
Since the beginning of the European Union, the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area \citep[as developed in:][]{Mundell:1961}, 
all the way to the very survival of the euro in light of the budgetary problems of some of its member
states. The effects of monetary unions on a number of macroeconomic indicators,
with inflation being the most important one in this context, is in the center of
an ongoing debate. This concerns the issue of short--run and steady--state
inflation uncertainty -- in context to inflation expectations -- dealt with in
\citet{cap}, or the degree of similarity of short--run dynamic properties of the inflation rates in EU
countries, which is the topic of an investigation by \citet{palomba}.

@@

Basically, the way towards the EMU basically consisted of three stages: stage~I
(1990--1994) as a phase of liberalization, stage~II (1995--1998) a phase of
convergence, and stage~III (1999--2001) the transition period, which ended with
the introduction of the euro as legal tender. A possible fourth stage is the
continuing integration of new member states, mostly former communist countries.
\readme{problem of breaks addressed} The possible influence of these different stages on the inflation rate dynamics
suggested itself for a more thorough investigation.

To inquire into possible changes in the dynamics of the inflation rate following
these stages, which may have led to considerable alterations, seems a promising task.

@@

%Overall, we find four periods during which structural breaks
%happened. Two of them in the 1990s (early and late), another one in the early 2000s, and the last period covers the financial crisis. There is also a 
%tendency towards a lower mean inflation rate, combined with a decreasing difference between
%individual inflation rates.

\readme{more about aggregation...} Overall, we find that the countries of the core euro--area (as defined in Section~\ref{sec:results}) did not change after the second stage and ended up with a low mean and variance. The southern euro--area member states and Ireland decreased their inflation rates almost to a level of the core countries but with a higher variance and a non--symmetric skewness. The Eastern European countries did reduce mean and variance significantly, especially the two countries that joined the euro--zone first. In the case of the Nordic non--euro countries, there is either no change at all or a clear decline in mean and variance after overcoming national crisis. Generally, inflation rates, which were very different in the beginning of the sample, are becoming more similar in terms of all three moments.

@@

@@

To explore the question of convergence, we proceed in two steps:
First, in Section~\ref{sec:bpresults}, the estimation technique proposed
in Section~\ref{sec:model} is applied to each of the countries included in
the data set (see Section~\ref{sec:data}). Second, Section~\ref{sec:aggregate}
aggregates the results from each individual country to reveal different
potential convergence patterns over time within three groups of countries:
core-euro (\Sexpr{paste(colnames(hicps)[euro], collapse = ", ")}),
PIIGS  (\Sexpr{paste(colnames(hicps)[piigs], collapse = ", ")}), and the
other countries (\Sexpr{paste(colnames(hicps)[others], collapse = ", ")}).

@@

@@

Finally, the inflation rates for the whole euro area\footnote{This is the euro--area 17 countries series as provided by the \citet{oecd} and covers the total euro--area. This series is calculated as a weighted average of the HICPs of now seventeen euro countries and starts in January 1996.}, are considered and three different regimes are found: (1)~until 
1999 with both a low mean and variance, (2)~the third and final stage of monetary integration,
accompanied with a significant increase in the mean, (3)~a period encompassing
the overheating of the economy that culminated in the financial crisis, showing a big increase
in variance and a lower mean inflation.

\readme{perhaps skip?} One clearly visible trend is the rise in inflation volatility in some of the euro countries (Austria, Greece, Italy, Luxembourg and Spain) and the  series for the whole euro area. \citet{emerson} emphasize  the fact that a high inflation rate is also more
variable and uncertain and thus causes more relative price variability, leading
to a less efficient price mechanism. From an economic point of view, a very volatile inflation rate will 
likely  contribute to greater macroeconomic instability. In the literature, there is no consensus about the effects of higher 
inflation volatility on the inflation rate. \citet{grier} find no evidence 
that higher inflation uncertainty raises the average inflation rate -- at least in the case of the USA. However,
\cite{jaro} finds a positive correlation between the level and the standard deviation of inflation. Naturally, an
efficient price mechanism will benefit from inflation rates that do not diverge too much from the expectations.

\fixme{some concluding remarks about the EU level bridging the current gap to the next subsection}
In any case, the EU series as a whole -- which is already an aggregate -- does sum up the general trend, namely the low inflation rates during the late 90s, a higher level until the financial crisis and a lower mean inflation rate thereafter. However, this series is too highly aggregated and starts too late to be able to draw precise conclusions regarding the convergence issue.

@@

\clearpage
\subsection{Convergence on an aggregate level}\label{sec:aggregate}

@@

In order to both aggregate the data to a number of sub--groups but avoid too high a level of compression, as is the case of the EU series, 
the countries are are grouped into the ``core'' of the eurozone, the PIIGS countries  and all the others. For this, we took account of possible
breaks in the moments. These groups are chosen so as to reflect the general dynamics of European inflation rates during the time period of interest.

The result of this is presented in Table~\ref{tab:results2} that shows lower quartile,
median, and upper quartile (25\%, 50\%, 75\% percentiles) of the first three moments of the
inflation series in January \Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]} and \Sexpr{yrs[4]}.

\fixme{also ``zoom out'' more, a big picture needs to be established}
The time periods are chosen according to four fundamental change periods. Namely, the early 1990s, with breaks in
Austria, Finland, France, Sweden, and UK. The late 1990s, with breaks in Greece, Italy, Luxembourg, Portugal, Czech Republic,
Estonia, and Hungary. The period after the second phase of convergence, with breaks in the early 2000s (Poland, Slovakia, and Slovenia)
and the last period consisting of two countries with the most visible effect of the financial crisis (Ireland and Spain).

In the core--zone, there is a decrease in the lower quartile from \Sexpr{yrs[1]} to
\Sexpr{yrs[2]}. However, the interquartile range decreased by almost 20\%. The variance increased
slightly, but also with a 15\% decrease in interquartile range. The skewness
stayed symmetric in the median, with almost an 8\% increase in the interquartile range.
\readme{overview}Generally, there are no dramatical changes in the core--zone and after the initial stages, none of the core--countries
had another break. This reflects the fact that their respective monetary policies were already closely connected.

The PIIGS countries significantly decreased the median inflation rates from \Sexpr{yrs[1]} to \Sexpr{yrs[4]}. 
Whereas the interquartile range was very tight in \Sexpr{yrs[1]}, it increased in \Sexpr{yrs[2]} and again in \Sexpr{yrs[4]}.
However, the median values are much smaller. The variance increased from \Sexpr{yrs[1]} to \Sexpr{yrs[2]} but did not change much 
in \Sexpr{yrs[4]}. The interquartile range did not change much too. Median skewness changed considerable 
from highly positive to negative values. 
\readme{overview, skewness effects} The decline in skewness and the narrowing of the interquartile range reflects the
fact that significant shocks to the inflation rates, which for some series were quite common in the past, have become something of a rarity. 
Still, the PIIGS countries tend to come closer towards the core--countries on behalf of the general trend of lower values and 
narrower bands.

The other countries were successful in decreasing all three moments from \Sexpr{yrs[1]} to \Sexpr{yrs[3]}. Still, their median
inflation rate is higher than those of the core countries. Also, the interquartile range of the mean decreased considerably. 
Contrary to the euro--zone and the PIIGS countries, variance decreased also, both in terms of the median value and the
interquartile range. The skewness decreased as well, as did the interquartile range of the skewness parameter.
\readme{overview} The other countries started with higher mean inflation values -- since some of them are Eastern European countries -- and 
effectually decreased the median inflation to a level much more similar to the core--zone than in \Sexpr{yrs[1]}. The bulk of changes 
in these countries took effect after \Sexpr{yrs[2]}, as a result of both an ending of the transitionary phase (from planned to free market economies) and the intended entry
into the EMU.

In context of the convergence issue, we find that, with the interquartile range of the mean of all the aggregated zones declining and getting tighter, there is  
evidence of a convergence towards some common mean inflation rate in the countries of our sample, both within the subgroups and the three groups put together.
Regarding the variance, the PIIGS countries, and the core countries as well, saw an increase over time, which is mostly due to the influence of the financial
crisis with its ``up--down--up'' turn (as depicted in Figure~\ref{fig:glmean}) leading to a higher variance.

@@

@@

More specifically, the mean inflation of the euro area  from a GL fit in the 
year before the crisis (\Sexpr{yearmon(2007 + 7/12)}--\Sexpr{yearmon(2008 + 6/12)}) was
$\Sexpr{round3(glmean(window(eu, start = yearmon(2007 + 7/12), end = yearmon(2008 + 6/12))))}$
and after the crisis (\Sexpr{yearmon(2008 + 7/12)}--\Sexpr{yearmon(2009 + 6/12)}) it dropped to
$\Sexpr{round3(glmean(window(eu, start = yearmon(2008 + 7/12), end = yearmon(2009 + 6/12))))}$
before rising again in the following year  (\Sexpr{yearmon(2009 + 7/12)}--\Sexpr{yearmon(2010 + 6/12)}):
$\Sexpr{round3(glmean(window(eu, start = yearmon(2009 + 7/12), end = yearmon(2010 + 6/12))))}$.
Another way to bring out this pattern in a graphical way is
Figure~\ref{fig:glmean}. It depicts the inflation rates (in gray) and a 12-month rolling mean (in red),
calculated again as the first moment of a fitted GL distribution.
The strong increase in mean inflation during the boom years of 2007/2008
accompanied by an equally strong deflation following the burst of the bubble
is clearly visible for most countries. However, the level of mean inflation
did not change persistently, as the inflation rates seem to have returned to
their prior path, conveying the same message as the inflation series of the
euro area.

@@

To assess changes in the dynamics of the inflation rates of a number of European countries both within
and without the EMU, a new method for estimating structural breaks resting on the 
assumption that the inflation rates are GL distributed has been  developed.

The results suggest that a GL distribution is appropriate for the modeling of the seasonally 
adjusted inflation rate series and that failing to account for a change in the skewness of the 
distribution might lead to potentially misleading conclusions concerning the dynamics. 
The decline in size and frequency of mostly positive outliers in many inflation rate series 
would have gone unnoticed without the skewness parameter to capture these changes.

At the beginning of the sample, many countries did have very high inflation rates, especially the Eastern and Southern countries. 
For the core--zone, convergence was already achieved in \Sexpr{yrs[2]}. The PIIGS countries lowered their inflation rates considerably in preparation 
to the adoption of the euro. Over the time --  however -- many inflation rates declined considerably and are now much more in line with the inflation rates of the traditional low inflation countries of the core--euro zone. This is reflected especially well in the Eastern European countries, that lowered their high initial inflation level in the late 90s or prior to the entry into the ERM~II. They clearly benefited from the monetary discipline that is demanded for those countries who want to join the EMU.

@@

\begin{figure}[htp!]
<<slovenia-plots1, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
plot(x$data, main = "", ylab = "Monthly inflation rate (in percent)", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
if(x$nbreaks > 0) lines(confint(x$breakpoints, breaks = x$nbreaks))
plot(x$efp, functional = supLM(0.1), main = "", ylab = "LM statistics")
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{\label{fig:slovenia1} Inflation rate series for Slovenia
with breakpoint estimate (\Sexpr{breakdates(x$breakpoints)}), associated confidence interval, and fitted mean
from generalized logistic distribution (left). Corresponding sup$\mathit{LM}$ test (right)
with sequence of $\mathit{LM}$~statistics,
critical value at 5\% level (horizontal red line line), and estimated breakpoint (vertical dashed line).}
\end{figure}

\begin{figure}[t!]
<<slovenia-plots2, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
for(i in 1:2) plot(x$model[[i]], moments = TRUE,
  xlim = c(-1.5, 2.2), ylim = c(0, 1.3),
  breaks = seq(-1.5, 2, by = 0.25),
  xlab = paste(names(x$model)[i], "\n", "Goodness-of-fit p-value: ",
    format.pval(summary(x$model[[i]])$chisq.test$p.value, digits = 4),
    sep = ""))
@
\caption{\label{fig:slovenia2} Histogramm of observed monthly inflation rates in
Slovenia for \Sexpr{names(x$model)[1]} (left) and \Sexpr{names(x$model)[1]} (right),
along with fitted generalized logistic probability density function, associated
moments, and $\chi^2$ goodness-of-fit test.}
\end{figure}

\clearpage
\begin{figure}[t!]
\begin{center}
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  gray_red1 <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0))
  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
    xlab = "Time", ylab = "Monthly inflation rates")
  for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (gray) and estimated means from
12-month rolling GL fits (red) over time for all countries.}
\end{center}
\end{figure}

@@

\begin{figure}[htp!]
\begin{center}
<<graphs-nochange, fig=TRUE, height=6.3, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(2, 4, 8, 14)))
@
\caption{\label{fig:nochange} No change.}
\end{center}
\end{figure}

\begin{figure}[p!]
\begin{center}
<<graphs-phase1, fig=TRUE, height=4.9, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(1, 6, 7)))
@
\caption{\label{fig:phase1} Phase~1 convergence.}
\end{center}
\end{figure}

\begin{figure}[p!]
\begin{center}
<<graphs-phase2, fig=TRUE, height=6.3, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(9, 12, 13, 16)))
@
\caption{\label{fig:phase2} Phase~2 convergence.}
\end{center}
\end{figure}

\clearpage
\begin{figure}[p!]
\begin{center}
<<graphs-crisis, fig=TRUE, height=3.5, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(11, 19)))
@
\caption{\label{fig:crisis} Financial crisis.}
\end{center}
\end{figure}

\clearpage
\begin{figure}[p!]
\begin{center}
<<graphs-noneuro, fig=TRUE, height=3.5, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(20, 21)))
@
\caption{\label{fig:noneuro} Non-euro.}
\end{center}
\end{figure}

\clearpage
\begin{figure}[p!]
\begin{center}
<<graphs-eastern, fig=TRUE, height=9.1, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(3, 5, 10, 15, 17, 18)))
@
\caption{\label{fig:eastern} Eastern countries.}
\end{center}
\end{figure}

\clearpage
\section{Tables}

\begin{table}[t!]
\begin{center}
\footnotesize
\begin{tabular}{llrrrlll}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline

\clearpage
\begin{table}[t!]
\begin{center}
\begin{tabular}{llrrrr}
\hline
Moment & Quartile & \Sexpr{yrs[1]} & \Sexpr{yrs[2]} & \Sexpr{yrs[3]} & \Sexpr{yrs[4]}\\
\hline
